---
title: "methylation level of sample ABLG 47592 teting data"
author: "charles-gumbi"
date: "2025-06-18"
output:
  html_document:
    df_print: paged
---

# Load necessary libraries

```{r libraries}
library(ggplot2)
library(dplyr)
library(ggplot2)
library(dplyr)
library(readr)
```

# Read the CpG methylation file

```{r read_data}
# Read the file into a data frame
CpG_df <- read.delim("~/Sebastes/Sebastes_test_data/data/CpG_metylation_levels_bedgraph.txt", header = FALSE, comment.char = "t")
# names on the bed file 
#col 1. The chromosome/contig/scaffold name
#col 2. The start coordinate
#col 3. The end coordinate
#col 4. The methylation percentage rounded to an integer
#col 5. The number of alignments/pairs reporting methylated bases
#col 6. The number of alignments/pairs reporting unmethylated bases

# Assign column names (based on your earlier description)
colnames(CpG_df) <- c("chrom", "start", "end", "methylation_percentage", "methylated_count", "unmethylated_count")
# View the first few rows
head(CpG_df)
```

# Editing dataframe 

```{r editing_data}
#Remove rows with missing values 
CpG_df <- na.omit(CpG_df)
# Remove "genome_hic_" and "_RagTag" from the 'chrom' column
CpG_df$chrom <- gsub("genome_hic_", "", CpG_df$chrom)
CpG_df$chrom <- gsub("_RagTag", "", CpG_df$chrom)
```

# Plot methylation percentage by chromosome

```{r methylation_by_chromosome}
# Plot methylation percentage as a scatter plot
cpg_plot <- ggplot(CpG_df, aes(x = start, y = methylation_percentage, 
              color = chrom)) + 
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ chrom, scales = "free_x") +
  theme_minimal() +
  labs(title = "Scatter plot for CpG methylation by scaffold",
       x = "Genomic position",y = "Methylation percentage (%)") +
  theme(legend.position = "none")

# Save CgG plot
ggsave("~/Sebastes/Sebastes_test_data/figures/CpG_methylation_plot.png", plot = cpg_plot, width = 10, height = 6, dpi = 300)

```

# Cytosine followed by any other base and guinine

```{r CHG}
# Read the CHG file
CHG_df <- read.delim("~/Sebastes/Sebastes_test_data/data/CHG_methylation_levels_bedgraph.txt", header = FALSE, comment.char = "t")
# Assign column names
colnames(CHG_df) <- c("chrom", "start", "end", "methylation_percentage", "methylated_count", "unmethylated_count")

# Clean chromosome names
CHG_df$chrom <- gsub("genome_hic_", "", CHG_df$chrom)
CHG_df$chrom <- gsub("_RagTag", "", CHG_df$chrom)

# Remove rows with missing or out-of-range methylation values
CHG_df <- CHG_df %>%
  filter(!is.na(methylation_percentage) & methylation_percentage >= 0 & methylation_percentage <= 100)

# Create scatter plot
chg_plot<- ggplot(CHG_df, aes(x = start, y = methylation_percentage, color = chrom)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ chrom, scales = "free_x") +
  theme_minimal() +
  labs(title = "Scatter plot of CHG methylation by chromosome",
       x = "Genomic position",
       y = "Methylation percentage (%)") +
  theme(legend.position = "none")
# Save CHH plot
ggsave("~/Sebastes/Sebastes_test_data/figures/CHG_methylation_plot.png", plot = chg_plot, width = 10, height = 6, dpi = 300)
```

# Cytosine followed by two other bases but not a guinine

```{r CHH}
# Read the CHG file
CHH_df <- read.delim("~/Sebastes/Sebastes_test_data/data/CHH_methylation_levels_bedgraph.txt", header = FALSE, comment.char = "t")

# Assign column names
colnames(CHH_df) <- c("chrom", "start", "end", "methylation_percentage", "methylated_count", "unmethylated_count")

# Clean chromosome names
CHH_df$chrom <- gsub("genome_hic_", "", CHH_df$chrom)
CHH_df$chrom <- gsub("_RagTag", "", CHH_df$chrom)

# Remove rows with missing or out-of-range methylation values
CHH_df <- CHH_df %>%
  filter(!is.na(methylation_percentage) & methylation_percentage >= 0 & methylation_percentage <= 100)

# Create scatter plot
chh_plot <- ggplot(CHH_df, aes(x = start, y = methylation_percentage, color = chrom)) +
  geom_point(alpha = 0.6, size = 1) +
  facet_wrap(~ chrom, scales = "free_x") +
  theme_minimal() +
  labs(title = "Scatter plot of CHH methylation by chromosome",
       x = "Genomic position",
       y = "Methylation percentage (%)") +
  theme(legend.position = "none")

# Save CHH plot
ggsave("~/Sebastes/Sebastes_test_data/figures/CHH_methylation_plot.png", plot = chh_plot, width = 10, height = 6, dpi = 300)

```


# CpG methylation across the genome 

```{r cpg_genome}
# Create scatter plot
chg_plot <- ggplot(CHG_df, aes(x = start, y = methylation_percentage)) +
  geom_point(alpha = 0.6, size = 1) +
  #facet_wrap(~ chrom, scales = "free_x") +
  theme_minimal() +
  labs(title = "Scatter plot of CHG methylation by chromosome",
       x = "Genomic position",
       y = "Methylation percentage (%)") +
  theme(legend.position = "none")
# Save CHH plot

```


# Methylation bias on controls 

```{r mbias_data}
# Load the data
mbias_data <- read_tsv("~/Sebastes/Sebastes_test_data/data/ABLG_47592_controls_combined_mbias.tsv")
```

# Sebsetting puc19 (positive control) 

```{r subsetting_puc19_data}
# Subset the data
pcu19 <- subset(mbias_data, chr == "plasmid_puc19c" & context == "CpG")

# Calculate methylation percentage
pcu19 <- pcu19 %>%
  mutate(methPerc = 100 * nMethylated / (nMethylated + nUnmethylated))

# Average methylation
mean(pcu19$methPerc, na.rm = TRUE)
#[1] 38.42835

# Plot pcu19 data
pcu19_plot <- ggplot(pcu19, aes(x = Position, y = methPerc, color = context, linetype = strand)) +
    geom_line() +
    labs(
      title = paste("Mbias plot for pcu19-positve control"),
      x = "Read Position",
      y = "Methylation perc (%)") +
    theme_minimal()
  
# Save puc19 mbias plot
ggsave("~/Sebastes/Sebastes_test_data/figures/pcu19_mbias.png", plot = pcu19_plot, width = 10, height = 6, dpi = 300)

```

# Sebsetting lamda control (negative control)

```{r subseting_lambda_data}
#Subset the data
lambda <- subset(mbias_data, chr == "phage_lambda" & context == "CpG")

# Calculate methylation percentage
lambda <- lambda %>%
  mutate(methPerc = 1 * nMethylated / (nMethylated + nUnmethylated))

# Average methylation
mean(lambda$methPerc, na.rm = TRUE)
#[1]0.4150602

# Plot lambda data
lambda_plot <- ggplot(lambda, aes(x = Position, y = methPerc, color = context, linetype = strand)) +
    geom_line() +
    labs(
      title = paste("Mbias Plot for lambda-negetive control"),
      x = "Read Position",
      y = "Methylation perc (%)"
    ) +
    theme_minimal()
  
# Save lanbda mbias plot
ggsave("~/Sebastes/Sebastes_test_data/figures/lambda_mbias.png", plot = lambda_plot, width = 10, height = 6, dpi = 300)
```



